Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-06-09 paper/letter.tex
narrative

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MG$^3$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
features in its Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
We show that these special features, a symmetric triple peaked
emission line, could be explained by two different models: an
homogeneous sphere undergoing gas bulk rotation or a clumpy outflowing
multiphase inter-stellar medium.
Using archival observational information about \tol\ we 
find that both models require a total dynamical mass in the range
$2.5\times 10^{9}$M$_{\odot}$ $<M_D<$  $4.5\times 10^{9}$M$_{\odot}$.  
This dynamical mass is at least $15\pm 5$ times
larger than the neutral mass hydrogen.
We argue that a possibility to explain the excess in
dynamical mass is the presence of a super-massive black hole.}

@@

The Ly-$\alpha$ emission line is not exclusive of distant galaxies. 
Any galaxy with low dust content and ongoing star formation has the
right conditions to show this line.  
There are, for instance,  local Universe surveys that target
Ly-$\alpha$ emission in nearby dwarf star forming galaxies. 
\cite{LARS}. 
The study of nearby LAE samples has allowed the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

@@

The most important consequence of bulk rotation is that, even for a
spherical gas distribution, the line morphology now depends on the
viewing angle respect to the rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity. 
When the rotational velocity is close to the half-line width of the
static line the line becomes single peaked, as it is observed in
\tol, a unique feature that other theoretical models find
difficult to reproduce without invoking complex gas distributions.

In the case of a multiphase interstellar medium there are a wide variety
of possible emerging spectra, mostly due to the increased complexity
of a model that must take into account the behaviour of neutral hydrogen
clumps dispersed in a tenous intraclump medium. 
This variety includes single, double and triple peaked
spectra. 
A clumpy interstellar medium also facilitates the escape of Ly
continuum and \lya\ photons.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two important
features. 
First, it has symmetric profile  without any systemic displacement
from the expected line-center. 
Second, it shows a triple peaked profile. 
Commonly, the Ly-$\alpha$ line features a single or asymmetric double peak. 
These two special features in \tol\ had evaded a natural physical
interpretation so far.

We find that a rotating spherical gas distribution provides a
satisfactory explanation of \tol's puzzling features.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol with the
overplot from our best fit model from the full radiative transfer
simulation.   
The parameters for the best fit are a rotational velocity of
$v_{max}=300$\kms, an optical depth $\tau=1\times10^7$ and temperature
of $T=1.5\times 10^{4}$K.
The optical depth and temperature can be translated into a column
density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Our model is also able to constrain the angle
between the rotation axis and the observational line-of-sight to
$\theta=65^{\circ}\pm$. 
To determine the uncertainties we perform a Markov Chain Monte
Carlo calculation using an approximate analytic solution for the
Ly-$\alpha$ spectrum including rotational effects.

Radio surveys of the 21cm line have put an upper limit to the neutral
hydrogen mass in \tol\ of $M<2.65\times 10^{8}$ M$_{\odot}$. 
Together with our velocity and column density results we constrain the
total dynamical mass of the HI region to be in the range $2.1\times
10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times 10^{9}$M$_{\odot}$ and the
diameter of HI region should have a diameter of $0.11 <
D/\mathrm{kpc}<0.34$.  
This makes the dynamical mass  at least $7$ to $25$ times larger than
the neutral Hydrogen mass.

We consider three possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass can be constrained by measurements of the 
stellar continuum, this can only represent $xx$M$_{\odot}$ of the total
contribution. 
Considering a standard dark matter halo, it can only contribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

A future observational test of this possibility in \tol\ would require
integral field unit measurements spatially resolving its spatial
extent. 
\tol\ spans a region of $4$ arcseconds. 
An instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.

Why is triple peaked emission not common in Lyman-alpha emitting
gaalaxies?   
The ubiquity of galaxy outflows overimposed to the rotation
transforms the line into the common double or single peaked
line.
Therefore, the absence of HI outflows in \tol\ is another special
feature that permitted a symmetric triple peaked lya line to emerge.

The implications of our finding for the study of LAEs, including the
first generation of galaxies are manifold. 
First of all, it demonstrates the importance of including rotation as
theoretical feature to interpret observed spectra. 
Secondly, it provides a motivation to explore under what conditions
rotation dominated dynamics dominate over outflow
dominated dynamics. 
Finally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it would provide a new information to test and probe
theories on the co-evolution of galaxies and black holes in the first
generation of galaxies.

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MG$^3$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
We explain these features by two different models: an
homogeneous sphere undergoing gas bulk rotation and an interstellar
medium composed by clumps with random motions.
It is the first time that an observed \lya\ spectrum can be explained
with these physical conditions.
We  find that both models independently require high velocities that
translate into a dynamical mass at least $10$ times larger than the
neutral mass hydrogen in the galaxy. 
We argue that a possibility to explain the high
dynamical mass is the presence of a super-massive black hole.
The implications of our findings for the study of LAEs, including the
first generation of galaxies are manifold. 
It demonstrates the importance of considering rotation and multiphase
physics under the possible conditions shaping the \lya\ spectra of the
first galaxies. 
Aditionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black holes. 
}

@@

The Ly-$\alpha$ emission line is not exclusive of distant galaxies. 
Any galaxy with low dust content and ongoing star formation has the
right conditions to show this line.  
There are, for instance,  local Universe surveys that target
Ly-$\alpha$ emission in nearby dwarf star forming galaxies 
\cite{LARS}. 
The study of nearby LAE samples has allowed the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

@@

The most important consequence of bulk rotation is that, even for a
spherical gas distribution, the line morphology now depends on the
viewing angle respect to the rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity. 
When the rotational velocity is close to the half-line width of the
static line the line becomes single peaked and symmetric, a unique
feature that other theoretical models find difficult to reproduce
without introducing more complexity into the gas distribution.

This is the case of a multiphase interstellar medium.
The model is more complex as it must take into account the behaviour
of neutral dense hydrogen clumps dispersed in a tenous intraclump medium.  
Corresponding to the elaborate physical conditions in the model,
there are a wide variety of possible emerging spectra.
This variety includes single, double and triple peaked
spectra with different degrees of symmetry around the line's center.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: it is symmetric and single peaked.
Commonly, the Ly-$\alpha$ line has a single or asymmetric double peak. 
These two special features in \tol\ had evaded a physical
interpretation so far, because bulk rotation and a multiphase ISM can
independently explain the \lya\ line in \tol.

Figure 1. summarizes our findings.
Dots represent the observational data for \tol with the
overplot from our best fit models from the full radiative transfer
simulations for a rotating sphere (continous line) and a clumpy ISM
(dashed line).

The best parameters in the rotation model are a rotational velocity of 
$v_{max}=300^{+xx}_{-xx}$\kms, an optical depth $\log\tau=7^{+xx}_{-xx}$ and a temperature
of $T=1.5^{+xx}_{-xx}\times 10^{4}$K. 
This translates into a column density of $\log N_{HI} /
\mathrm{atoms\ cm}^{-2} =  20.5^{+xx}_{-xx}$.  
This model is also able to constrain the angle between the rotation
axis and the observational line-of-sight to
$\theta=65^{+xx}_{-xx}$$^{\circ}$.

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=71^{+17}_{-25}$ \kms,
the clumps ouflowing velocity $v_{\infty, \rm{cl}}=79^{+167}_{60}$
\kms and the fraction of the \lya emission that is comming comming from the cold clumps
$P_{cl}=0.72^{+0.20}_{-0.32}$. 
The muliphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $3.0^{+1.5}_{-1.8}\times
10^9$ M$_{\odot}$.

Radio surveys of the 21cm line have put an upper limit to the neutral
hydrogen mass in \tol\ of $M<2.65\times 10^{8}$ M$_{\odot}$ \cite{pustilnikmartin07}. 
In the case of the rotation model, this information help us to constrain the
diameter of the HI region where the \lya\ emission and transfer takes
place. 
We find this size to be in the range $0.11 < D/\mathrm{kpc}<0.34$, one
order of magnitude smaller than \tol's size in the optical.   
That size and the rotational velocity of $300$\kms put a constraint on
the dynamical mass of $4.5^{+2.1}_{-2.4} \times 10^{9}$M$_{\odot}$.

This is evidence for a dynamical mass at least 11 (multiphase case) to
17 (rotation case) times higher than the mass in neutral hydrogen.  
There are three main possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass in old stars hass been constrained by non-detections
in the K-band to be $<XXX$M$_{\odot}$.
Considering a standard dark matter halo, it can only contribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

A future observational test to clarify the physical nature of
\tol\ would require integral field unit measurements spatially
resolving its spatial extent. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
Another observational test includes the measurement of the escape
fraction of Ly continuum ionizing radiation. 
In the rotational model this fraction should be zero, while
the multiphase model predicts that in this case it should be around
$0.5^{+1.0}_{-0.4}$\%.

All in all, the mere existence of a strong LAE galaxy with a broad,
symmetric line is interesting.
It raises the question whether some high redshift LAEs have asymmetric
lines because the blue half was truncated by the intergalactic medium.
In this case the \lya\ radiation could emerge as a low surface
brightness glow, which may be connected to \lya\ halos, while also
influencing the way LAEs can be used as a probe of reionization.

These findings demonstrate the importance of including rotation and multiphase
conditions as features to model the \lya\ line in high redshift
galaxies.
Aditionally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it could correspond to a so far undetected black hole
in a dwarf galaxy, providing a new way to test and probe
theories on the co-evolution of galaxies and black holes in the first
generation of galaxies.
